Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-22 paper.Rmd
Redo the confusion matrices in LaTeX

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set. This appears to be an easily corrected coding error on the part of B&S.  ANDY ARE YOU SURE THIS PARAGRAPH IS CORRECT?  DON'T YOU WANT TO WEIGHT THE TEST BY THE TRAINING?

@@

B&S report (in their Table 4) confusion matrices for their forecasts for the first half of 2016 (2016-H1). To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%, i.e., an order of magnitude smaller. The positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data. This suggests that the forecasts were erroneously assessed using civil war incidence, not onset. By examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e., including ongoing civil wars, rather than civil war onset years only.

```{r, message=FALSE}
tab4o <- read_csv("data/table4-original.csv")
tab4f <- read_csv("data/table4-fixed.csv")

@@

ANDY, THIS SECTION 4.6 AND THE TABLE 4 ARE VERY CONFUSING. CAN YOU TRY TO CLEAN IT UP?
The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall value is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces two civil war onsets. Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.^[Another, minor issue is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.]

Predicting rare events like civil war onset is exceedingly difficult, and it is not unusual at all that most of the predictions are false positives. This will by definition be the case when there are no onsets in a time period, as is true with the default "Assuming Persistence" coding that is in the data. Correctly predicting the two onsets in the alternative "Assuming Change" coding is at face value a credit to the model's accuracy.

@@

@@

The intention with the "Weighted by PITF" model is to combine the PITF test predictions with the Escalation test predictions, i.e. weight the latter with the former. Due to an error, the PITF _training_ set predictions are used, not the _test_ set predictions.

See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly, in the 6-month version of this file.

>>>>

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set. This appears to be an easily corrected coding error on the part of B&S.

@@

B&S made forecasts for the first half of 2016 (2016-H1), and assess the accuracy of these forecasts. The forecasts themselves are probabilistic, i.e. range from 0 to 1, but for the purpose of scoring the forecasts they consider the 30 highest predictions to have been positive forecasts ("1") and the remaining forecasts to have been for no onset ("0"). They then construct confusion matrices for the forecasts, in their Table 4 and which we replicate here in the top half of our Table \ref{tab:table4}. They use two alternative codings of the outcomes for 2016-H1, one of which they call "Assuming Persistence" (this is what is in their replication data), and one of which they call "Assuming Change" (hand-coded in one of their replication scripts).

With the "Assuming Persistence" version of the outcomes, shown in the first confusion matrix there are 15 civil war onsets, of which the forecasts capture 13 (true positives) and miss 2 (false negatives). With the "Assuming Persistence" coding of outcomes, shown in the second confusion matrix, there are 16 onsets and the forecasts capture 14 of those with, still, 2 misses. Fifteen to 16 civil war onsets in the first half of 2016 is incredible, especially considering that the entire historical data from 2001 to 2015 only has 20 recorded civil war onsets. Indeed, it turns out that B&S scored their forecasts for civil war \textit{onset} using civil war \textit{incidence}, i.e. counting ongoing civil wars as if they were continuing onsets.

```{r, include=FALSE}
# AB 2020-09-22: I redid the confusion matrices in latex so that the resulting table is clearer than what I could easily do with kable() programatically here in this chunk. 
tab4o <- read_csv("data/table4-original.csv")
tab4f <- read_csv("data/table4-fixed.csv")

@@

\begin{table}
\caption{Replication of B\&S Table 4: Confusion Matrices for Six-month Escalation Model forecasts for the first half of 2016.}\label{tab:table4}
\centering
\begin{tabular}{ll|c|c|l}
\multicolumn{5}{c}{\textbf{Original confusion matrices with civil war \textit{incidence}}} \\
% BS, assuming persistence
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 132 & 17 \\
\cline{3-4}
& $Y=1$ & 2 & 13 \\
\cline{3-4}
% BS, assuming change
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 132 & 16 \\
\cline{3-4}
& $Y=1$ & 2 & 14 \\
\cline{3-4}
\\ \midrule
\multicolumn{5}{c}{\textbf{Fixed confusion matrices with civil war \textit{onset}}} \\
% fixed, assuming persistence
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Persistence}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 134 & 30 \\
\cline{3-4}
& $Y=1$ & 0 & 0 \\
\cline{3-4}
% fixed, assuming change
\multicolumn{1}{l}{\multirow{4}{*}{Assuming Change}} & \multicolumn{3}{c}{} \\
\multicolumn{2}{c}{}& \multicolumn{1}{c}{$P=0$} & \multicolumn{1}{c}{$P=1$} \\
\cline{3-4}
& $Y=0$ & 134 & 28 \\
\cline{3-4}
& $Y=1$ & 0 & 2 \\
\cline{3-4}
\end{tabular}
\end{table}

The correct confusion matrices that use civil war onsets to score the forecasts are shown in the second part of Table \ref{tab:table4}. In the "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. The alternative "Assuming Persistence" coding records 2 civil war onsets in 2016-H1. As a result, if we compare the precision and recall of the forecasts with the correct scoring to the incorrect scoring in B&S, almost all of the forecasts are false positives. This is the norm when predicting rare events, and thus not unusual at all. The forecasts do correctly capture the 2 civil war onsets in the "Assuming Change" version of the outcome coding. This is, at face value, a credit to the model's accuracy.

@@

Conversely, not all factors that are "theoretically" important are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @wgb:2007 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes.

@@

We believe the intention with the "Weighted by PITF" model is to combine the PITF test predictions with the Escalation test predictions, i.e. weight the latter with the former. Due to an error, the PITF _training_ set predictions are used, not the _test_ set predictions.

See `1mo_run_escalation_weighted_PITF.R` line 4, where the PITF predictions are taken from the training data set (`train$pred_prob_plus1`). The next line is a hack extending the shorter `weight` vector with missing values to avoid a R warning when it is multiplied with the longer vector of escalation model test set predictions. Similarly in the 6-month version of this file.
